---
title: "Reverse Engineering Project"
author: "Alex Taylor, Jordan Fine, Devin Etta"
date: "Date here"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story, [STORY NAME](STORY URL)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(janitor)
library(lubridate)

```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

```{r}
# Load required data
Donations <- read_csv("data/sports-political-donations.csv")



# Clean required data and prepare for analysis if needed. 
sports_donations <- read_csv("data/sports-political-donations.csv")
sports_donations <- sports_donations |>
  clean_names()
sports_donations <- sports_donations %>%
  mutate(amount = as.numeric(gsub("[$,]", "", amount)))


```


## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1 - Jordan

* **Sentence text**: “American professional sports owners have contributed nearly $47 million in federal elections since 2015… $34.2 million (72.9 percent) [went] to Republican campaigns or super PACs… compared to $10.1 million (21.5 percent) to Democrats.”

* **Analysis summary**: I think this quote includes an interesting statistic that provides the main point of the article. There is a big difference in how much money went to republicans vs democrats, this could mean something important or make for an important story. 


```{r}
# First, I am making two new data tables that will show me democratic and republican donors separately. I want to see who is donating the most, to which team, and in which party. 

republican_donors <- sports_donations |>
  filter(party == "Republican")

democrat_donors <- sports_donations |>
  filter(party == "Democrat")
#Now I have two more datasets that show me who is donating and which party they are in.I notice that there are significantly more observations in the republican dataset. This makes sense because significantly more people donated. 

#I am now writing code to find out how much money each party received in total, and even the average donation amount, I can also look at how many donors gave to each party.

sports_donations |>
  group_by(party) |>
  summarize(
    total_donated = sum(amount, na.rm = TRUE),
    average_donation = mean(amount, na.rm = TRUE),
    number_of_donors = n()
  ) |>
  arrange(desc(total_donated))
#This data is telling me the average donation amounts for each political party, that is pretty cool! The data isn't all that surprising, seeing as Republicans donated the most. I do find it interesting that the number of doners doesn't exactly mirror the total amount donated. 

```

### Sentence 2 - Alex

* **Sentence text**: Overall, 51.7 percent of all contributions by WNBA owners went to Democrats, with 42.3 percent going to Republicans.

* **Analysis summary**: I was not able to confirm these results. In my analysis, 76.6% of WNBA contributions went to Democrats, while 19.7% went to Republicans. At first, I thought this was because an owner owned a team in a league other than the WNBA, but after this ended up not being the case. 

```{r}
# Put code to reverse engineer sentence here

# Filter to only focus on WNBA players, then see how many contributions were made to each party
WNBA_Contribs <- Donations |>
  filter(League == "WNBA") |>
  group_by(Party) |>
  summarise(numb_contrib = n())

# Find the percentage of contributions each party got
WNBA_Contribs <- WNBA_Contribs |>
  mutate(
    pct_contrib = numb_contrib / sum(numb_contrib) * 100
  )

# Display results of code below this codeblock
WNBA_Contribs


```

### Sentence 3 - Devin (Did 5th sentance in document)

* **Sentence text**: “Forty owners account for more than $40 million — more than 85 percent of all contributions. And looking specifically at these big spenders, it’s clear where their allegiances lie: 77.7 percent of that $40 million went to Republican causes or candidates, with 18.8 percent going to Democrats.” 
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

#The data needs to be clean like this because the amount data set needs to be set to numeric
Clean_Donations <- Donations |>
  clean_names() |>
  mutate(amount = as.numeric(gsub("[$,]","",amount))) 

Clean_Donations



#top 40 donations code  
top_40donations <- Clean_Donations |>
  group_by(owner) |>
  summarize(total_amount = sum(amount))|>
  top_n(40,total_amount) |>
  summarize(total_from40 = sum(total_amount))

top_40donations
   #calculate the total contributions       
 total_donation_amount <- sum(Clean_Donations$amount)
   
 #percentage 
 percent_top40 <- (top_40donations$total_from40/ total_donation_amount) * 100

# Display results of code below this codeblock
  
 #the percent of the top 40 owners 
 percent_top40

```


### Sentence 4 - Alex

* **Sentence text**: Of the 160 owners and commissioners in ESPN’s database, 74 contributed a majority of their publicly searchable funds to Republican campaigns, with 48 donating a majority to Democrats and the rest sending most of their money to causes that aren’t fully aligned with either party.

* **Analysis summary**: I was able to confirm that 48 owners donated the majority of their funds to Democrats, however, I was not able to confirm that 74 owners contributed the majority of their funds to Republican campaigns. This is because, in my analysis, 70 owners contributed the majority of their funding to Republicans. This four owner difference could be, in part, because the number of owners in the dataset we have is 158, while the article stated that it's datebase has 160 owners in it. 

```{r}
# Put code to reverse engineer sentence here

# Turn the amount column into a numeric type, instead of a character
Donations_AmountClean <- Donations |>
  mutate(Amount = str_replace_all(Amount, "\\$", ""),
         Amount = str_replace_all(Amount, "\\,", ""),
         Amount = as.numeric(Amount)
         )
# Create a dataframe that lists the owners, which party they donatated to, and how much they donated to that party
Owners_PartyDons <- Donations_AmountClean |>
  group_by(Owner, Party) |>
  summarise(Contrib = sum(Amount))

# Add a column with the total amount each owner donated

      # Create a dataframe with the total amount each owner donated
      Owner_Total <- Donations_AmountClean |>
        group_by(Owner) |>
        summarise(Owner_Total = sum(Amount))
      
      # Join that dataframe to the first dataframe (that has the list of owners and party donations)
      Owners_PartyDons <- Owners_PartyDons |>
        left_join(Owner_Total)

# Create a column listing what percent each owner donated to each party

Owners_PartyDons <- Owners_PartyDons |>
  mutate(
    Pct_Don = (Contrib / Owner_Total) * 100
  )

# Find out how many owners contributed the majority of their funds to Republican campaigns

Majority_Rep <- Owners_PartyDons |>
  filter(Party == "Republican" & Pct_Don > 50)

# Find out how many owners contributed the majority of their funds to Democratic campaigns

Majority_Dem <- Owners_PartyDons |>
  filter(Party == "Democrat" & Pct_Don > 50)

# Display results of code below this codeblock

nrow(Majority_Rep)
nrow(Majority_Dem)
nrow(
  Donations |>
    group_by(Owner) |>
    summarise(count = n())
)

```

### Sentence 5

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

-30-