---
title: "Reverse Engineering Project"
author: "Alex Taylor, Jordan Fine, Devin Etta"
date: "Date here"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story, [STORY NAME](STORY URL)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(janitor)
library(lubridate)

```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

```{r}
# Load required data
Donations <- read_csv("data/sports-political-donations.csv")



# Clean required data and prepare for analysis if needed. 


```


## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1 - Jordan

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 2 - Alex

* **Sentence text**: Overall, 51.7 percent of all contributions by WNBA owners went to Democrats, with 42.3 percent going to Republicans.

* **Analysis summary**: I was not able to confirm these results. In my analysis, 76.6% of WNBA contributions went to Democrats, while 19.7% went to Republicans. At first, I thought this was because an owner owned a team in a league other than the WNBA, but after this ended up not being the case. 

```{r}
# Put code to reverse engineer sentence here

# Filter to only focus on WNBA players, then see how many contributions were made to each party
WNBA_Contribs <- Donations |>
  filter(League == "WNBA") |>
  group_by(Party) |>
  summarise(numb_contrib = n())

# Find the percentage of contributions each party got
WNBA_Contribs <- WNBA_Contribs |>
  mutate(
    pct_contrib = numb_contrib / sum(numb_contrib) * 100
  )

# Display results of code below this codeblock
WNBA_Contribs

```

### Sentence 3

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 4

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 5

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

-30-