---
title: "lab_03"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)

# Load janitor
library(janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab that are in the data folder: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows
payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=369008)

# Load in the Maryland grants and loan data
md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv", guess_max=19482)
```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system.

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

A1: I found 98 records. 

```{r}
payments <- payments |> 
  clean_names() |>
  mutate(agency_name = str_to_upper(agency_name), vendor_name = str_to_upper(vendor_name))

pg_schools <- payments |> 
  filter(str_detect(vendor_name, "PRINCE GEORGES") | str_detect(vendor_name, "PRINCE GEORGE'S") | str_detect(vendor_name, "PRINCE GEORGE") | str_detect(vendor_name, "PG"), str_detect(vendor_name, "SCHOOL") | str_detect(vendor_name, "EDUCATION")
         )
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2: The 2024 fiscal year had the largest total amount of payments. One result that stands out to me is that there is no data for 2023. Maybe this is an error on my part, or maybe this is an avenue for further reporting. Another interesting result is that, broadly speaking, spending increases as the years increase. 

```{r}
# Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year.

pg_schools |>
  group_by(fiscal_year) |> 
  summarise(total_pay = sum(amount)) |> 
  arrange(desc(total_pay))

```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received.

Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe in words how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3. To only have one row per vendor you will need to standardize the name used by the vendor so that all instances of it have the same name.

Here are vendors that have similar names (and are probably the same company):

A MORTON THOMAS & ASSOC INC		
A MORTON THOMAS & ASSOCIATES INC

AEROSOL MONITORING & ANALYSIS INC			
AERSOL MONITORING & ANALYSIS INC

AMERICAN DESIGN ASSOCIATE	
AMERICAN DESIGN ASSOCIATES		
AMERICAN DESIGN ASSOCIATES INC

To make these vendors be identical, in most cases we would have to truncate part of the name. For example, we would have to remove "IATES" from "A MORTON THOMAS & ASSOCIATES INC". We would also have to change the spelling of some vendors, like in the second vendor listed above, which has two different spellings for "Aerosol". To do this, we would have to use a function like mutate() that allows us to modify the data frame. We would most likely also have to use function like str_detect() so that we do not have to manually change each row. 


```{r}
# Finding vendors
UMD_CapitalProjects <- payments |> 
  filter(str_detect(agency_name, "MARYLAND") | str_detect(agency_name, "MD"),
         str_detect(agency_name, "UNIVERSITY"),
         str_detect(agency_name, "CAPITAL") | str_detect(agency_name, "PROJECT")
           )

# List vendors and total amount received
UMD_CapitalProjects |> 
  group_by(vendor_name) |> 
  summarise(amt_received = sum(amount))

```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which column(s) contain the information you want, your code should find just the grants for stem cell research in a single filter. Then calculate the total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4: There was a spike in stem cell research during 2009 (the first year we have data for), then it decreased in 2010 and stayed relatively the same (with some fluctuation) until 2023, when it spiked again. 

```{r}
# Clean data
md_grants_loans <- md_grants_loans |>
  clean_names() |>
  mutate(grantor = str_to_upper(grantor), grantee = str_to_upper(grantee), description = str_to_upper(description))

# Find stem cell grants and total amount
md_grants_loans |> 
  filter(str_detect(description, "STEM CELL")) |>
  group_by(fiscal_year) |> 
  summarise(total_amount = sum(amount))
  

```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5: According to my analysis, there are 18 individual listed grantees, though some of them, especially those referring to the university, have different names. So, in reality, there are probably less than 18 individual grantees. To clean these results, we would have to standardize the spelling of each grantee. For example, UMD is spelled three different ways, so we would need to modify them to match each other. The grantees that interest me the most are Ovaltz, Zymetis and Leukosight, which are, at least according to Google, are located inside UMD's campus, but do not seem to be affiliated with the university. I'd like to know why that is, and what ties those companies have to the university. 

```{r}
md_grants_loans |> 
  filter(str_detect(zip_code, "20742")) |>
  group_by(grantee) |> 
  summarise(total_amount = sum(amount)) |> 
  arrange(desc(total_amount))

```
