---
title: "lab_05"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Tabula

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse, plus any other packages you will need to clean data and work with dates.
library(tidyverse)
library(janitor)
library(lubridate)

```

## Get Our PDF

We'll be working with the [911 overdose calls from Baltimore County](https://drive.google.com/file/d/1qkYuojGF_6WKFr5aNQxmewDzcKyOiJFr/view?usp=share_link). You'll want to download it to a place you'll remember (like your Downloads folder, or the labs folder in your repository). The goal is to extract the tables within it, export that to a CSV file, load it into RStudio and ask some questions.

## Extract Data from PDF Using Tabula

Start Tabula, then go to <http://127.0.0.1:8080/> in your browser. Click the "Browse" button and find the PDF file and click "open", and then click the "Import button" in Tabula. This will take a few seconds or longer.

This PDF has a single table spread over multiple pages to extract. We're going to make a single dataframe from this table, exporting it to a CSV file that you will load into R. In Tabula, highlight the table and click the "Preview & Export Extracted Data" button. You may want to play with including or excluding the column headers - YOU SHOULD HAVE FIVE COLUMNS OF DATA.

Save the CSV (it should be called `tabula-Baltimore County; Carey, Samantha log OD.csv` by default) to your lab_05/data folder.

From there, you will need to read in the data, and add or fix headers if necessary. You can choose to include the headers from the PDF in your exported CSV files OR to exclude them and add them when importing. `read_csv` allows us to do this ([and more](https://readr.tidyverse.org/reference/read_delim.html)).

## Load and clean up the data in R

You will need to read in and clean up the data so that it can be used for analysis. By "clean" I mean the column headers should not contain spaces and they should have meaningful names, not "x1" or something similar. How you do that is up to you, but you can use select() with or without the minus sign to include or exclude certain columns. You also can use the `rename` function to, well, rename columns. Importantly, you'll need to ensure that any columns containing a date actually have a date datatype. Our friend `lubridate` can help with this.

```{r}
# load data
BaltCo_Overdoses <- read_csv("tabula-Baltimore County; Carey, Samantha log OD.csv", col_names = FALSE)

# rename columns and clean
BaltCo_Overdoses <- BaltCo_Overdoses |>
  clean_names() |> 
  rename(date = x1, time = x2, case_nbr = x3, evtyp = x4, loc = x5) |>
  mutate(date = mdy(date))


```

## Answer questions

Q1. Write code to generate the number of calls that occurred on each date. Which date in 2022 had the most overdose calls, and how many? Look at the total number of rows in your result and explore the range of dates - based on your result, do you believe there are any days with no overdose calls at all? Explain why or why not.

A1. Both July 7th and October 10th had the most overdose calls. They both had 23 calls. I do believe that there are days with zero overdose calls. This is because I only have 329 rows in this tibble, but there are 365 days in the year.

```{r}
BaltCo_Overdoses |>
  filter(str_detect(date, "2022")) |> 
  group_by(date) |>
  summarise(call_total = n()) |>
  arrange(desc(call_total))

```

Q2. You want to understand if there's a pattern in the day of the week that overdose calls are made. Add a column to your dataframe that displays what day of the week each date represents. You should search for how to do that using lubridate. Then write code to calculate the number of calls for each day of the week, and add a column to that result that calculates the percentage of all calls that occurred on each day of the week (so you want a dataframe with the day of the week, total number of calls and the percentage of calls on that day out of the total number of all calls). Describe your findings to me.

A2. According to my analysis, the most number of overdoses happened on Saturdays, with Sundays being a close second. However, the percentage of overdoses per day was pretty evenly split, with the lowest percent only being 12.7% Based on this, I would not say there is a definitive pattern to which days of the week have the most overdose calls. 

```{r}
# Add day of the week
BaltCo_Overdoses <- BaltCo_Overdoses |>
  mutate(
    day = wday(date, label = TRUE)
  )

# group by day and find percentage
BaltCo_Overdoses |>
  group_by(day) |>
  summarise(count = n()) |>
  mutate(
    pct = count/sum(count) * 100
  ) |>
  arrange(desc(pct))

```

Q3. Now let's look at locations. Which ones have the most calls? How would you describe them (feel free to search for more information on them)? Is there anything about the structure of the original data that might make you less confident in the counts by location or date?

A3. Of the top four locations, two of them (4540 SILVER SPRING RD and 330 LEEANNE RD) are just random houses, while the other two (PC 02 and PC 06) are police precincts. One thing that makes me feel less confident in the location counts is that some of the cells have two pieces of data in them. For example, the entries for PC 02 and PC 06 also include both the address and the PC description. So, if there was another cell that had one or the other, the count would be off. The only thing that seems off about the date data is that the original year is in listed as 22 or 23, and not the full year: 2022 or 2023. 

```{r}
BaltCo_Overdoses |>
  group_by(loc) |>
  summarise(count = n()) |>
  arrange(desc(count))

```

Q4. What's the best story idea or question you've seen as a result of the work you've done in this lab?

A4. The best story idea that I've seen is that, of the top four overdose call locations, two of them are from seeming random houses in the county. What's the story of these houses and why did so many overdose calls come from them? Was there some sort of organization there, or was it just someone with a bad drug habit? 
