---
title: "lab_01"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your class reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the UMD courses data.

**Task** Run the codeblock below to load the data, which is contained in an RDS file.

```{r}
umd_courses <- read_rds("lab_01.rds")
```

## Answer questions

**Q1** Which department has the highest mean (average) number of seats for its classes? You need to show the department with the highest average first. Which one is the most surprising/interesting to you? **A1**

```{r}
umd_courses |> 
  group_by(department) |> 
  summarise(
    avg_seat = mean(seats)
  ) |> 
  arrange(desc(avg_seat))

# I'm surprised that the First-Year Innovation & Research Experience (FIRE) courses are in the top five. I didn't think there were that many students admitted into the program (and thus need that many seats).
```

**Q2** Find out the greatest number of times any individual class has been offered in our dataset. Your code must show the greatest number of classes first.
**A2**

```{r}
umd_courses |> 
  group_by(id) |> 
  summarise(
    course_count = n()
  ) |>
  arrange(desc(course_count))

#29
```

**Q3** Count the number of courses for each instructor and department. Which instructor has taught the most?
**A3**

```{r}
umd_courses |> 
  group_by(instructors)|> 
  summarise(
    times_taught = n()
  ) |> 
    arrange(desc(times_taught))

# Jason Kuo has taught the most, though more courses have not had an instructor assigned (Instructor: TBA).

```

**Q4** Given the results for Q3, what would have made it easier to find the answer? You don't have to know *how* to accomplish this; I want you to describe in words what would make the answer easier to see. How confident are you that the answer you got is the right one, and why? **A4**

It would have been easier to find the answer if we were able to see the different number of both terms and individual courses each instructor taught, that way we would know if the instructors have taught for a long time, or just taught a large quantity of courses. I'm not 100% confident in my answer because we don't know if there were instructors who changed their name, or if their name was misspelled at some point. In other words, our calculations are only as good as the data we have, which could be influenced by human error. 

**Q5** Take one of the questions you listed in the final question of the pre_lab and try to answer it here in code. If you don't know how to fully answer it, write what code you can and describe in words what else you would need to do. If you got an answer, is it interesting? **A5**

I can't answer any of my questions with the data that we have. All of them require obtaining additional data. But, here's how I would go about answering one of those questions:

To find out how many courses are general education courses, I would first need to find and add data to the dataset that listed the general education credit of each course. Then, I would group by and count the number of courses that have each general education requirement. Then, I would find the sum of all of those classes, which should give me the total amount of general education courses offered.
